Almost orthogonal vectors

### 2014-12-10 project3/paper/paper.tex
Edits to paper

\maketitle
\tableofcontents

\section{Introduction}
Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
In this paper, we consider the problem of minimizing $\epsilon$ for various $n$ and $N$.

For $n=N$, it is clear that the minimum value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis. For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have 
\[\epsilon_1\leq \epsilon_2.\] 
Otherwise, given a configuration of $N+1$ vectors in $\mathbb R^n$ with $\epsilon=\epsilon_2$, we could simply remove one vector and obtain a configuration of $N$ vectors whose $\epsilon$ would be less than or equal to $\epsilon_1$.

@@

In this paper, we shall see that for $n=2$, the minimum value of $\epsilon$ will be given by 
\[\epsilon=\cos^2\pi/N.\] 
We will also derive a lower bound on $\epsilon$ for any choice of $n$ and $N$. In addition, we shall see that, for $N=n+1$, the minimum value of $\epsilon$ is given by
\[\epsilon=\frac{1}{n^2},\] 
and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex. We shall also exhibit an optimal configuration of six vectors in $\mathbb R^3$, with corresponding $\epsilon$ given by
\[\epsilon=\frac 1 5.\]

@@

However, we see that the problem gets harder for larger $N$. Let us first consider the specific case of $N=3$; once we build some intuition for the problem, we will try to generalize to generic $N$.
\\

@@

Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $1$. Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$. Hence our theorem statement is equivalent to proving that $\alpha$ is equal to $\cos^2 \theta_n$.

@@

Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$. (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)

@@

Given this lemma, we now prove the above theorem.

@@

Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin.

@@

Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$. Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$. Figure $1$ shows this optimum configuration.

@@

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ not in $\{j, j+1, \ldots, k-1\}$.

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by the above Lemma.

@@

We can use a well-known lemma to provide a useful lower bound on $\epsilon$. Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.

@@

\section{Some Optimal Configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{n=3,N=6}
In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are four possible dot products we get from these six vectors: $1,ab,-ab,a^2-b^2$. This gives us three possible values of $\epsilon$: $1,(a^2-b^2)^2,(ab)^2$. Let us consider the solution we get when $a^2-b^2=ab$. If we solve the resulting system according to the constraint $a^2+b^2=1$, we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we have that $\epsilon=\frac{1}{5}$, implying that these six points give us an optimal configuration.
\subsection{N=n+1}
Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we will find a configuration of vectors that achieves this lower bound. First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.

@@

Before proceeding with the theorem, we notice that when $n=2$, the configuration of $N=3$ vectors that produced the minimum value of $\epsilon$ made up a simplex centered at the origin.

@@

Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs. We describe this method in greater detail below.

@@

At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector.

To make this more precise, we present Python code for the algorithm above.

@@

We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$. Before proving an upper bound on what the optimum value of $\epsilon$ can be, we introduce the parallelogram law.

@@

\subsection{Variations: Minimizing the dot product}
Let us consider the situation in which we try to minimize $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for $n+1$ vectors in an $n$ dimensional space.

@@

Since $\delta_{min} \le -\frac{1}{n}$ and $\delta_{min} \ge -\frac{1}{n}$, we can conclude that $\delta_{min}=-\frac{1}{n}$. From this, we can see that a $\epsilon = \frac{1}{n^2}$ is achievable for the general case of when $N=n+1$ -- this leads us to conclude that the minimum possible $\epsilon$ value for $N=n+1$ is at most $\frac{1}{n^2}$.
\end{proof}

Note that proving that the minimum possible epsilon for the case of $N = n+1$ is equal to $1/n^2$ is much harder, and we leave this for future work.

>>>>

\maketitle

\section{Introduction}
Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
In this paper, we consider the problem of minimizing $\epsilon$ for different values of $n$ and $N$.

For $n=N$, it is clear that the minimum value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis. For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have 
\[\epsilon_1\leq \epsilon_2.\]

Otherwise, given a configuration of $N+1$ vectors in $\mathbb R^n$ with $\epsilon=\epsilon_2$, we could simply remove one vector and obtain a configuration of $N$ vectors whose $\epsilon$ would be less than or equal to $\epsilon_1$.

@@

In this paper, we shall see that for $n=2$, the minimum value of $\epsilon$ will be given by 
\[\epsilon=\cos^2\pi/N.\]

We will also derive a lower bound on $\epsilon$ for any choice of $n$ and $N$. In addition, we shall see that, for $N=n+1$, the minimum value of $\epsilon$ is given by
\[\epsilon=\frac{1}{n^2},\] 
and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex. We shall also exhibit an optimal configuration of six vectors in $\mathbb R^3$; this optimal configuration produces an $\epsilon$ equal to
\[\epsilon=\frac 1 5.\]

@@

However, we see that the problem gets harder for larger $N$. Let us first consider the specific case of $N=3$; once we build some intuition for the problem, we will try to generalize to arbitrary $N$.
\\

@@

Our argument hinges on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $1$.

Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$. Hence our theorem statement is equivalent to proving that $\alpha$ is equal to $\cos^2 \theta_n$.

@@

Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ which is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$. (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)

@@

Given this lemma, we now prove the theorem stated above.

@@

Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin (multiplied by $-1$).

@@

Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$. Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$. Figure $2$ shows this optimum configuration.

@@

Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$. Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$. Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.

From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.

@@

We can use a well-known lemma to provide a useful lower bound on $\epsilon$ for arbitrary $(n, N)$ pairs. Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.

@@

\section{Some Optimal Configurations}
Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
\subsection{$n=3,N=6$}
In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$ from the above theorem. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
$$
(a,b,0),(a,0,b),(b,0,a)
$$
$$
(a,-b,0),(a,0,-b),(-b,0,a)
$$
There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$. This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$. It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$. If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
$$
a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
$$
If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$; coupling this knowledge with the fact that for $(n, N) = (3, 6)$, we see that these six points give us an optimal configuration.
\subsection{$N=n+1$}
Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space. From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so if we show that there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$, then we can conclude that $1/n^2$ is the optimal value of $\epsilon$ for $N = n+1$.

To do this, we first define what we mean by an $n$-dimensional simplex.

@@

Before proceeding with the theorem, we notice that when $n=2$, the configuration of $N=3$ vectors that produced the minimum value of $\epsilon$ made up a simplex centered at the origin (an equilateral triangle).

@@

Since visualizing in spaces of dimensionality $3$ or greater is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs. We describe this method in greater detail below.

@@

At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector. We present pseudocode for this algorithm above.

@@

We now consider a variation of the above problem, where we actually try to place $N$ vectors in $n$-space as far apart as possible from each other. More concretely, if we define $\delta$ as

$$\delta = \max_{i \neq j} v_i \cdot v_j$$,

then we can formalize our problem as finding the $N$ vectors $v_1, v_2, \ldots, v_N$ that minimize $\delta$.

Before proving what the optimum value of $\delta$ is for $N= n+1$, we introduce the parallelogram law.

@@

\subsection{Variations: Minimizing the dot product}
Given the above bound, we proceed to state and prove the following theorem.

@@

Since $\delta_{min} \le -\frac{1}{n}$ and $\delta_{min} \ge -\frac{1}{n}$, we can conclude that $\delta_{min}=-\frac{1}{n}$.
\end{proof}
